Modular analysis of gene expression data with R

### 2009-10-15 isanote/modular.tex
isanote updated, seems to be ready.

\title[Modular analysis]{Modular analysis of gene expression data with
  GNU R}
\author[G\'abor Cs\'ardi \textit{et~al}]{G\'abor Cs\'ardi\,,$^{1,2}$,
  Zolt\'an Kutalik\,$^{1,2}$ and Sven Bergmann\,$^{1,2}$}
\address{$^{1}$Department of Medical Genetics, and 
  $^{2}$Swiss Institute of Bioinformatics,
  University of Lausanne, Rue de Bugnon 27, CH-1005 Lausanne,
  Switzerland.}

@@

\begin{abstract}
TODO
\section{Summary:} TODO
\section{Availability:}
\href{http://www.unil.ch/cbg/homepage/software.html}%
{http://www.unil.ch/cbg/homepage/software.html}
\section{Contact:} \href{Sven.Bergmann@unil.ch}{Sven.Bergmann@unil.ch}
\end{abstract}

@@

% What is biclustering.

Biclustering is the idea of clustering the rows and the columns of a
data matrix together, usually not independently. In other words, the
goal of a biclustering algorithm is to find blocks in the reordered
data matrix that have, such that the blocks have correlated rows and
columns. It is important to note that these blocks (or biclusters,
modules) have context, i.e. the rows only need to be correlated across
the columns of the block, but not across other columns of the data
matrix; and vice-versa.

When it comes to gene expression data, a bicluster means a subset of
genes ($G$) (or probes, probe sets) and a subset of samples ($S$),
such that genes $G$ are exactly coexpressed across $S$ and samples $S$
have correlated expression profile exactly across $G$.

% What is the ISA.

The Iterative Signature Algorithm (ISA), \cite{sa,isa,isamod}, is a 
powerful biclustering algorithm. It is able to identify biclusters
that overlap and it is resilient to noise.

\section{Methods}

\begin{figure*}
\centering
\includegraphics[width=0.9\textwidth]{isa2workflow3}
\caption{Work flow of a typical modular analysis study, with the
  \Rpackage{eisa} package (left); and four bicluster visualization
  techniques (right). The input of the \Rpackage{eisa} pipeline is a
  BioConductor \Rclass{ExpressionSet} object. This is normalized to
  $Z$-scores (1); the ISA iteration is performed on it using smart or
  random seeds (2); and the biclusters are then merged and filtered
  based on the robustness measure (3). The \Rpackage{eisa} package
  contains tools to perform many GO, KEGG, etc. enrichment calculation
  quickly (4), and several bicluster visualization tools (5).
  Subfigures \subfig{A}, \subfig{B}, \subfig{C} and \subfig{D} were generated
  using the acute lymphoblastic leukemia data set, see
  \cite{chiaretti04} and the \Rpackage{ALL} R package.
  \subfig{(A)} A condition plot for a simple bicluster. Each sample is
  represented as a bar, its height is its score in the bicluster. 
  Samples above the red (none in this case) and samples
  below the green line are part of the biclusters. The coloring
  indicates cases and controls. This module can separate control and
  case samples.
  \subfig{(B)} A module tree. Each rectangle is a bicluster. See the
  definition of the module tree in the text. The modules are colored
  according to their Gene Ontology enrichment $p$-values, the codes of
  the enriched GO categories are shown in the top-left corner of the
  rectangles. The top-right corner shows the number of genes and
  conditions in the bicluster. The numbers on the horizontal axes are
  the gene thresholds used for finding the modules.
  \subfig{(C)} Heatmap for a single module, showing correlated
  expression and the genes and samples. The red lines are the gene and
  sample scores.
  \subfig{(D)} Profile plot, for a single module. On the top plot the
  expression of all samples are plotted, the samples of the module
  with black. The red and green lines show the mean of the two
  groups. The bottom plot is the same for the genes of the module.
}
\label{fig:workflow}
\end{figure*}

Let us first explain the briefly the steps of a typical modular
analysis study for gene expression data.

% Batch correction
\paragraph{Batch correction}
Often, the ISA is applied to samples from different experiments. In
this case it is crucial to address the question of experimental
variation. Several methods exist to solve this problem,
see e.g. \cite{johnson07} for an algorithm that has a GNU R
implementation.

% Filtering
\paragraph{Gene filtering}
Depending on the application, it can be advantageous to remove genes
that are not expressed in any of the samples, especially because the
ISA normalization (next step) tends to amplify their effect.

% Normalization. 
\paragraph{ISA normalization}
ISA is an iterative algorithm. In each step it computes weighted sums
of expression levels for a number of genes or a number of
samples. Since different genes typically show different levels of base
expression and variance, it is important to standardize expression
levels to $Z$-scores. The ISA uses two sets of $Z$-scores, one
calculated for each gene across all samples, the other for each sample
across all genes.

% Random and smart seeding.
\paragraph{Random and smart seeding}
The ISA used with random seeds is an unsupervised algorithm, it uses no
external knowledge to find the biclusters. ISA with smart seeds is
biased towards genes or pathways of interest, or sample annotation.
This latter form can be considered as a semi-supervised approach.

% The ISA iteration. 
\paragraph{The ISA iteration}
This step is essentially performing the ISA and finding the biclusters
from the starting seeds.

% Merging the modules.
\paragraph{Merging the modules}
It is possible that more seeds converge to the same, or very similar
biclusters. This step eliminates such duplicates.

% The robustness measure and filtering the modules. 
\paragraph{Robustness of biclusters}
To access the significance of a bicluster, we designed a robustness
measure, that can be used to filter out spurious modules. This is done
based on performing ISA on the scrambled input data.

% Module trees.
\paragraph{Module trees}
The ISA works with two stringency threshold parameters, the gene
threshold and the sample threshold. ISA modules can be organized into
a directed graph, a module tree, in which there is an edge from module
$A$ to module $B$, if the ISA converges to module $B$ from module $A$,
with the same threshold parameters  that were used to find module
$B$. A module tree provides a hierarchical modular description of a
data set.

\section{Implementation}

@@

TODO

\paragraph{Funding\textcolon} TODO

\paragraph{Conflict of interest\textcolon} None declared.

>>>>

\title[Modular analysis]{Modular analysis of gene expression data with R}
\author[G\'abor Cs\'ardi \textit{et~al}]{G\'abor Cs\'ardi\,,$^{1,2}$,
  Zolt\'an Kutalik\,$^{1,2}$ and Sven Bergmann\,$^{1,2}$}
\address{$^{1}$Department of Medical Genetics, and 
  $^{2}$Swiss Institute of Bioinformatics,
  University of Lausanne, Rue de Bugnon 27, CH-1005 Lausanne,
  Switzerland.}

@@

\begin{abstract}
\section{Summary:} 
Biclustering means clustering the rows and the columns of a
data matrix together, usually not independently. A bicluster is, then,
a block of the reordered input matrix. When it comes to gene
expression data, a bicluster means a subset of genes ($G$) (or probes,
probe sets) and a subset of samples ($S$), such that genes $G$ are
exactly coexpressed across $S$ and samples $S$ have correlated
expression profile exactly across $G$.

The Iterative Signature Algorithm (ISA), \cite{sa,isa}, is a 
powerful biclustering algorithm. It is able to identify biclusters
that overlap and it is resilient to noise.

\section{Availability:}
\href{http://www.unil.ch/cbg/homepage/software.html}%
{http://www.unil.ch/cbg/homepage/software.html}
\section{Contact:} \href{Sven.Bergmann@unil.ch}{Sven.Bergmann@unil.ch}
\end{abstract}

@@

% How the ISA works

The ISA iteration is started from an input seed, this is 
refined it by adding and/or removing more genes and/or
samples until the process converges to a stable bicluster. The
procedure is typically performed to a large number of seeds; these can
be either chosen randomly (random seeding, unsupervised method), or
based on the data (smart seeding, semi-supervised method).

The output of ISA is a set of potentially overlapping
biclusters. These biclusters are not binary, both genes and samples of
a bicluster have scores, normalized between minus one and one. Genes
and samples with scores further away from zero have a stronger
association to the module.

In this short note has two parts, in Section~\ref{sec:methods} we look
over the steps of a typical modular analysis, and in
Section~\ref{sec:implementation} introduce the R packages that
implement the ISA.

\section{Methods}%
\label{sec:methods}

\begin{figure*}
\centering
\includegraphics[width=\textwidth]{isa2workflow3}
\caption{Work flow of a typical modular analysis study, with the
  \Rpackage{eisa} package (left); and four bicluster visualization
  techniques (right). The input of the \Rpackage{eisa} pipeline is a
  BioConductor \Rclass{ExpressionSet} object. This is normalized to
  $Z$-scores (1); the ISA iteration is performed on it using smart or
  random seeds (2); and the biclusters are then merged and filtered
  based on the robustness measure (3). The \Rpackage{eisa} package
  contains tools to perform many GO, KEGG, etc. enrichment calculation
  quickly (4), and several bicluster visualization tools (5).
  Subfigures \subfig{A}, \subfig{B}, \subfig{C} and \subfig{D} were generated
  using the acute lymphoblastic leukemia data set, see
  \cite{chiaretti04} and the \Rpackage{ALL} R package.
  \subfig{(A)} A condition plot for a simple bicluster. Each sample is
  represented as a bar, its height is its score in the bicluster. 
  Samples above the red (none in this case) and samples
  below the green line are part of the biclusters. The coloring
  indicates cases and controls.
  \subfig{(B)} A module tree. Each rectangle is a bicluster; see the
  definition of the edges in the text. The modules are colored
  according to their Gene Ontology enrichment $p$-values, the codes of
  the enriched GO categories are shown in the top-left corner of the
  rectangles. The top-right corner shows the number of genes and
  conditions in the bicluster. The gene thresholds used for finding
  the modules is shown on the horizontal axes.
  \subfig{(C)} Heatmap for a single module, showing correlated
  expression for the genes and samples. The red lines are the gene and
  sample scores.
  \subfig{(D)} Profile plot, for a single module. On the top plot the
  expression of all samples is plotted, the samples of the module
  with black. The red and green lines show the mean of the two
  groups. The bottom plot is the same for the genes of the module.
}
\label{fig:workflow}
\end{figure*}

Let us first explain briefly the steps of a typical modular
analysis study for gene expression data.

% Batch correction
\emph{Batch correction.}
Often, the ISA is applied to samples from different experiments. In
this case it is crucial to address the question of experimental
variation. Several methods exist to solve this problem,
see e.g. \cite{johnson07} for an algorithm that has a GNU R
implementation.

% Filtering
\emph{Gene filtering.}
Depending on the application, it can be advantageous to remove genes
that are not expressed in any of the samples, especially because the
ISA normalization (next step) tends to amplify their effect.

% Normalization. 
\emph{ISA normalization.}
ISA is an iterative algorithm. In each step it computes weighted sums
of expression levels for a number of genes or a number of
samples. Since different genes typically show different levels of base
expression and variance, it is important to standardize expression
levels to $Z$-scores. The ISA uses two sets of $Z$-scores, one
calculated for each gene across all samples, the other for each sample
across all genes.

% Random and smart seeding.
\emph{Random and smart seeding.}
As mentioned previously, ISA can be performed with random or smart
seeds, based on the application.

% The ISA iteration. 
\emph{The ISA iteration.}
This step is essentially performing the ISA and finding the biclusters
from the starting seeds.

% Merging the modules.
\emph{Merging the modules.}
It is possible that more seeds converge to the same, or very similar
biclusters. This step eliminates such duplicates.

% The robustness measure and filtering the modules. 
\emph{Robustness of biclusters.}
To access the significance of a bicluster, we designed a robustness
measure, that can be used to filter out spurious modules. This is done
based on performing ISA on the scrambled input data.

% Module trees.
\emph{Module trees.}
The ISA works with two stringency threshold parameters, the gene
threshold and the sample threshold. ISA modules can be organized into
a directed graph, a module tree, in which there is an edge from module
$A$ to module $B$, if the ISA converges to module $B$ from module $A$,
with the same threshold parameters  that were used to find module
$B$. A module tree provides a hierarchical modular description of a
data set.

\section{Implementation}%
\label{sec:implementation}

@@

TODO
\paragraph{Funding\textcolon} TODO
\paragraph{Conflict of interest\textcolon} None declared.
